### 2017-01-19 response-to-reviewers.md

> Reviewer 1
> 
> This manuscript provides rules aimed at novice programmers to help
> them write more robust bioinformatics programs. The rules are all
> sensible, and biologists turned programmers (and some people with
> formal software engineering training) would do well to read them.
> 
> With any set of guidelines, and especially guidelines about
> programming, there is room for subjectivity and differences of
> opinion, so here are mine. The authors are free to include or exclude
> these comments in their revision.

>  Robustness
>
>The rules are framed in terms of increasing robustness.
>
> The piece starts out defining robustness in terms of portability,
> specifically outside the programmer's own environment - but then goes
> on to define it more broadly in terms of multiple criteria such as
> being kept under version control (which I don't think is a necessary
> condition for robustness - many robust programs were made before VCSs
> although I think using VCSs is excellent advice for a modern
> programmer)
> 
> The definition here is somewhat different from Wikipedia's, which defines it as:
> 
> "the ability of a computer system to cope with errors during execution[1][2] and cope with erroneous input."
> 
> I am not sure of the best name for the property the authors are aiming at, but it seems some combination of robustness, maintainability, quality, predictability - essentially better software.

Thanks for the feedback. We have clarified and condensed our definition of 'robust' software in the introduction, and have made version control a full rule.

### 2017-01-20 response-to-reviewers.md

Dear Editor and Reviewers,

Please find attached our revised paper for "Ten Simple Rules for Making Research Software More Robust". We have taken all reviewer comments into account in this revision and have produced a stronger paper. Each of the reviewer's concerns is addressed in-line below. Attached is also a version of the document with changes highlighted in orange.

Thank you and best regards,

Greg Wilson & Morgan Taschuk

@@

> Reviewer 1
> 
> This manuscript provides rules aimed at novice programmers to help
> them write more robust bioinformatics programs. The rules are all
> sensible, and biologists turned programmers (and some people with
> formal software engineering training) would do well to read them.
> 
> With any set of guidelines, and especially guidelines about
> programming, there is room for subjectivity and differences of
> opinion, so here are mine. The authors are free to include or exclude
> these comments in their revision.

>  Robustness
>
> The rules are framed in terms of increasing robustness.
>
> The piece starts out defining robustness in terms of portability,
> specifically outside the programmer's own environment - but then goes
> on to define it more broadly in terms of multiple criteria such as
> being kept under version control (which I don't think is a necessary
> condition for robustness - many robust programs were made before VCSs
> although I think using VCSs is excellent advice for a modern
> programmer)
> 
> The definition here is somewhat different from Wikipedia's, which defines it as:
> 
> "the ability of a computer system to cope with errors during execution[1][2] and cope with erroneous input."
> 
> I am not sure of the best name for the property the authors are aiming at, but it seems some combination of robustness, maintainability, quality, predictability - essentially better software.

@@

Thanks for the feedback. We have clarified and condensed our definition of 'robust' software in the introduction, and have made version control a full rule.

> Documentation and maintainability.
> 
> I find the relegation of documentation to an additional note rather
> than a rule surprising. It's also telling the authors don't call out
> maintainability as a desirable property (except in passing in the
> conclusions). If software is not thrown away, it frequently needs to
> be updated - e.g. to accomodate changes in input file types. This
> maintenance task often falls on people who are not the original
> author. If software does not have adequate documentation, it will be
> harder to maintain (either by the original author or by others).

We've promoted documentation to a full rule by merging the README and Usage rules into one and explicitly mentioning other types of documentation with appropriate references.

@@

>    Good software development practice
> 
> The rules do not mention anything about good software engineering
> practice, such as structured programming. While it's good to not be
> too specific and dictate specific fads or paradigms (and the intended
> audience is individual programmers not teams), no one would disagree
> that some kind of good structured programming practices are
> required to write programs where clarity, maintainability,
> extensibility and quality are of concern.
> 
> In general, the program should not repeat large pieces of code with
> the same logic - these should be abstracted into
> subroutines/functions/methods/classes etc. The program should not be a
> giant monolithic main() method. Behavior of subroutines should not be
> controlled through globals. etc.
> 
> While it's not necessary to replicate Knuth's three volumes, I think
> good software engineering practice deserves to be a rule, or at least
> called out.

Thanks for the feedback. We are focusing mostly on software that is preparing to be published or otherwise distributed, and so discussing the details of program flow and execution are out of scope. We do mention other good software engineering practices, such as source control, testing, and versioning as an introductory taste of software engineering.

@@

> Usage info
> 
> Good advice, but I wonder if 2 and 8 could be combined?

We consider usage to be part of the documentation, but a section of this rule referred to checking input parameters and echoing out parameters to the command line. These sections have been moved to the rules on controlling common operations and producing identical results.

> Versioning
> 
> All good advice.
> 
> As the authors note, GitHub makes this very easy. The advice is a bit succinct:
> 
> "as simple as adding an appropriate commit message or tag to version control"
> 
> As this article seems to be aimed as novice programmers who may not
> have been immersed in good, modern, open software development
> practice, it would be a good idea to include more details or pointers here.
> 
> In fact, I would recommend every programmer learns how to use a VCS
> and a hosted VCS like GitHub/GitLab/Bitbucket, to commit early and
> commit often, to use issue trackers, etc. Not just a matter of making
> public releases (which is important) but also as a matter of good
> practice even for software not intended to be released.

We think it's important too, but it's a tough sell in some crowds. Version control is its own rule now, and we've added a reference to another educational article, "A Quick Introduction to Version Control with Git and GitHub".

@@

> built utility / package manager
> 
> Excellent advice that is all too often ignored.
> 
> As noted above the presentation in combo with 4 is
> confusing. Generally a package manager is used when an external
> library, module or package is re-used. But as noted there is generally
> no advice on reusing these artefacts, as 4 is about running programs.

We believe we have addressed the discrepencies between these two rules with our re-write.

> Do not require root privileges
> 
> I certainly wouldn't contradict this but I wonder why it's called
> out. There are many possible inadvisable things we've all seen in years of programming, but maybe this could be combined with advice about running external programs (4) or making the software runnable outside a default user environment (7)

With the increasing torrent of 'big data', more and more calculations are outsourced into high performance compute clusters or cloud environments. When spawning jobs across tens to thousands of machines that may or may not share any storage, and likely only gives their users privileges in their own space, a requirement for needing root to install a package severely limits its reuse and distribution. Other people without administrative access to their machines are also cut off from using such software. Mostly, we've been bitten by this often enough to feel strongly that it should be its own rule.

@@

> In several cases, it seems that the paper might violate the first rule of “The 10 simple rules for writing a 10 simple rules paper” (http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1003858) with multiple rules combined into a single rule.

We reviewed all the sections you mentioned. In general, we believe that most of our *bolded* sections within each rule are refinements of the rule and not separate rules. We address each one of your concerns in-line below.

> For example rule 8 combines putting parameters on the command line with echoing all parameters to the log file to support reproducibility and provenance.

We've moved these recommendations around so that they fit better into their major rule than they did before.

@@

> Similarly rule 5 covers both automating the build process and automating the deployment and installation.

We've clarified this section to focus mostly on build automation and dependency acquisition.

@@

> Rule 4 also tends to stuff multiple rules (bolded) into one rule (( limit dependencies, ensure dependencies are available) these 2 are closely related), with “ use native functions” which is a very different idea and not what I would expect from a ‘reuse’ rule.

We removed this last 'native functions' rule and replaced it with a more appropriate discussion of robust dependencies.

@@

> Finally the ‘What we left out’ section implicitly defines an additional rule (documentation). In this particular case there could have been a single rule (Document your code and usage) that could encompass the points of both Rules 1 and 2 (Readme, usage) as well as the points from ‘What we left out’.

We combined all mentions of documentation into a single rule.

> In rule 8, I would tend to suggest generalizing the rule somewhat to suggest externalizing parameters rather than necessarily requiring them to be available as command line options. For example if a configuration file is used to provide all parameter values, this could accomplish the task of separating the parameters out from the code as well as providing a record of the values used for a run as they describe through echoing parameters to a log file.

This section has been re-written to take your concerns into account.

---

Thank you again for your time and effort.

### 2017-01-20 robust-software.tex

\newenvironment{reviewed}
{ \color{orange} }

@@

%% Include all macros below
\newcommand{\fixme}[2]{\textsc{\textbf{{#1}: {#2}}}}
\newcommand{\recommend}[1]{\textit{#1}}
\newcommand{\withurl}[2]{{#1}\footnote{\texttt{#2}}}
\newcommand{\rulemajor}[1]{\section{#1}}
\newcommand{\ruleminor}[1]{\textbf{#1}}

\begin{document}
\vspace*{0.2in}

@@

Being unable to
replicate results is so common that one publication refers to it as ``a rite of
passage''~\cite{baker2016}.
The root cause of this problem is that most research software
is essentially a prototype, and therefore is not
\emph{robust}. The lack of
robustness in published, distributed software
leads to duplicated efforts with little practical benefit,
which slows the pace of research~\cite{prabhu2011,lawlor2015}.
Not all software needs to be robust enough to run for other people,
but software that is published and distributed to the community should be.
Bioinformatics software repositories \cite{ison2016,brazas2012} catalogue dozens to
hundreds of tools that perform similar tasks:
for example,
in 2016 the Bioinformatics Links Directory included 84 different multiple sequence aligners, 141 tools
to analyze transcript expression, and 182 pathway and interaction resources.
Some of these tools are legitimate efforts to improve the state-of-the-art, but
often they are difficult to install and run~\cite{stajich2002,Seemann2013}, and are effectively abandoned
after publication~\cite{nekrutenko2012}.

@@

The good news is,
you don't need to be a professionally-trained programmer to write robust software.
In fact,
some of the best, most reliable pieces of software in many scientific
communities are written by researchers~\cite{prabhu2011,sanders2008}
who have adopted strong software
engineering approaches, have high standards of reproducibility, use good testing
practices, and foster strong user
bases through constantly evolving, clearly documented, useful, and useable software.
In the bioinformatics community, Bioconductor and Galaxy follow this
path~\cite{gentleman2004,afgan2016}.
Not all scientific software needs to be robust~\cite{varoquaux2015}, but if you
publish a paper about your software, it should, at minimum, satisfy these
rules.

So what \emph{is} ``robust'' software?
We implied above that it is software that works for people other than the original author,
on machines other than its creator's.
More specifically, we mean that:

@@

Our rules are generic and can be applied to all languages, libraries, packages,
documentation styles, and operating systems for both closed-source and open-source software. 
They are also necessary steps toward making computational research replicable and reproducible:
after all,
if your tools and libraries cannot be run by others,
they cannot be used to verify your results or as a stepping stone for future work~\cite{brown2013}.

\rulemajor{Rule 1: Use version control.}

@@

If you are new to version control,
it is simplest to treat it as ``a better \withurl{Dropbox}{http://dropbox.com}''
(or, if you are of a certain age, a better FTP),
and to use it simply to synchronize files between multiple developers and machines~\cite{blischak2016}.
Once you are comfortable working that way,
you should \ruleminor{use a feature branch workflow}:
designate one parallel copy (or ``branch'') of the repository as the master,
and create a new branch from it each time you want to fix a bug or add a new feature.
This allows work on independent changes to proceed in isolation;
once the work has been completed and tested,
it can be merged into the master branch for release.

\rulemajor{Rule 2: Document your code and usage}

@@

\item \textit{List a few example commands} to get a user started quickly.

@@

Documentation beyond the README and usage is up to the developer's discretion.
We think it is very important for developers to document their work, but our
experience is that people are unlikely do it during normal development
However, it
is worth noting that software that is widely used and contributed to has and
enforces the need for good documentation~\cite{gentleman2004}.

@@

In the spirit of code reuse and interoperability, developers often want
to reuse software written by others.
With a few lines, a call
is made out to another library or program and the results are incorporated into the
primary script. Using popular projects reduces the amount of code that
needs to be maintained and leverages the work done by the other software.

Unfortunately, reusing software (whether software libraries or separate executables)
introduces dependencies, which can
bring their own special pain. The interface between two software
packages can be a source of considerable frustration: all too
often, support requests descend into debugging errors produced by the
other project due to incompatible libraries, versions, or operating
systems~\cite{brown2013}. Even introducing libraries in the same programming
language can rely on software installed in the environment, and the problem
becomes much more difficult when relying on executables, or even on web
services.

@@

First,
\ruleminor{make sure that you really need the auxiliary program}. If you are
executing GNU sort instead of figuring out how to sort lists in Python,
it may not be worth the pain of integration. Reuse software that offers some
measurable improvement to your project.

Second, if launching an executable, \ruleminor{ensure the appropriate software and version is available}.
Either allow the user to
configure the exact path to the package, distribute the program with the
dependent software, or download it during installation using your
package manager. If the executable requires internet access, check for that
early in execution.

@@

Exercise caution especially when transitioning across languages or using
separate executables, as they tend to be especially sensitive to operating
systems, environments, and locales.

@@

To compile code, deploy applications, and automate other tasks,
programmers routinely use build tools like Make, Rake, Maven, Ant or MS Build.
These tools can also be used to manage runtime environments,
i.e.,
to check that the right versions of required packages are installed
and install or upgrade them if they are not.
As mentioned in Rule~5,
a package manager can mitigate some of the difficulties in software reuse.

@@

Every package should come with a small test script for users to run
after installation. Its purpose is not only to check that the software
is working correctly (although that is extremely helpful), but also to
ensure that it works at all. This test script can also serve as a
working example of how to run the software.

In order to be useful, \ruleminor{make the test script easy to find and run}.
Many build systems will also run unit tests if provided them at compile time.
If the build system is not amenable to testing, a
single file in the project's root directory named \texttt{runtests.sh}
or something equally obvious is a 
good solution that lets new users build their analysis from a working script.
For example, with its distribution, 
HISAT2 includes a full set of very small files,
and a 'Getting Started with HISAT2' section in its manual 
that leads you through the entire data lifecycle~\cite{pertea2016}.

@@

The usage message tells users what the program could do.
It is equally important for the program to tell users what it actually did.
Accordingly,
when the program starts, it should \ruleminor{echo all parameters and software
versions to standard out or a log file alongside the results}. This
feature supports greater reproducibility because any result can be
replicated with only the previous output files as reference.

>>>>

@@

%% Include all macros below
\newcommand{\fixme}[2]{\textsc{\textbf{{#1}: {#2}}}}
\newcommand{\recommend}[1]{\textit{#1}}
\newcommand{\withurl}[2]{{#1}\footnote{\texttt{#2}}}
\newcommand{\rulemajor}[1]{\section{#1}}
\newcommand{\ruleminor}[1]{\textbf{#1}}
\newcommand{\reviewed}[1]{{\color{black}#1}}
\begin{document}
\vspace*{0.2in}

@@

Being unable to
replicate results is so common that one publication refers to it as ``a rite of
passage''~\cite{baker2016}.
The root cause of this problem is that most research software
\reviewed{ is essentially a prototype, and therefore is not
\emph{robust}. The lack of
robustness in published, distributed software
leads to duplicated efforts with little practical benefit,}
which slows the pace of research~\cite{prabhu2011,lawlor2015}.
Bioinformatics software repositories \cite{ison2016,brazas2012} catalogue dozens to
hundreds of tools that perform similar tasks:
for example,
in 2016 the Bioinformatics Links Directory included 84 different multiple sequence aligners, 141 tools
to analyze transcript expression, and 182 pathway and interaction resources.
Some of these tools are legitimate efforts to improve the state-of-the-art, but
often they are difficult to install and run~\cite{stajich2002,Seemann2013}, and are effectively abandoned
after publication~\cite{nekrutenko2012}.

@@

The good news is,
you don't need to be a professionally-trained programmer to write robust software.
In fact,
some of the best, most reliable pieces of software in many scientific
communities are written by researchers~\cite{prabhu2011,sanders2008}
who have adopted strong software
engineering approaches, have high standards of reproducibility, use good testing
practices, and foster 
\reviewed{strong user
bases through constantly evolving, clearly documented, useful, and useable software.}
In the bioinformatics community, Bioconductor and Galaxy follow this
path~\cite{gentleman2004,afgan2016}.
\reviewed{Not all scientific software needs to be robust~\cite{varoquaux2015}, but if you
publish a paper about your software, it should, at minimum, satisfy these
rules.}

So what \emph{is} ``robust'' software?
\reviewed{We implied above that it is software that works for people other than the original author,
on machines other than its creator's.
More specifically, we mean that:

@@

Our rules are generic and can be applied to all languages, libraries, packages,
documentation styles, and operating systems for both closed-source and open-source software. 
They are also necessary steps toward making computational research replicable and reproducible:
after all,
if your tools and libraries cannot be run by others,
they cannot be used to verify your results or as a stepping stone for future work~\cite{brown2013}.}

\reviewed{\rulemajor{Rule 1: Use version control.}

@@

If you are new to version control,
it is simplest to treat it as ``a better \withurl{Dropbox}{http://dropbox.com}''
(or, if you are of a certain age, a better FTP),
and to use it simply to synchronize files between multiple developers and machines~\cite{blischak2016}.
Once you are comfortable working that way,
you should \ruleminor{use a feature branch workflow}:
designate one parallel copy (or ``branch'') of the repository as the master,
and create a new branch from it each time you want to fix a bug or add a new feature.
This allows work on independent changes to proceed in isolation;
once the work has been completed and tested,
it can be merged into the master branch for release.}

\reviewed{\rulemajor{Rule 2: Document your code and usage}

@@

\reviewed{\item \textit{List a few example commands} to get a user started quickly.}

@@

Documentation beyond the README and usage is up to the developer's discretion.
We think it is very important for developers to document their work, but our
experience is that people are unlikely do it during normal development
However, it
is worth noting that software that is widely used and contributed to has and
enforces the need for good documentation~\cite{gentleman2004}.}

@@

In the spirit of code reuse and interoperability, developers often want
to reuse software written by others.
With a few lines, a call
is made out to \reviewed{another library or program} and the results are incorporated into the
primary script. Using popular projects reduces the amount of code that
needs to be maintained and leverages the work done by the other software.

Unfortunately, reusing software \reviewed{(whether software libraries or separate executables)}
introduces dependencies, which can
bring their own special pain. The interface between two software
packages can be a source of considerable frustration: all too
often, support requests descend into debugging errors produced by the
other project 
\reviewed{due to incompatible libraries, versions, or operating
systems~\cite{brown2013}. Even introducing libraries in the same programming
language can rely on software installed in the environment, and the problem
becomes much more difficult when relying on executables, or even on web
services.}

@@

First,
\ruleminor{make sure that you really need the auxiliary program}. If you are
executing GNU sort instead of figuring out how to sort lists in Python,
it may not be worth the pain of integration. \reviewed{Reuse software that offers some
measurable improvement to your project.}

Second, \reviewed{if launching an executable,}
\ruleminor{ensure the appropriate software and version is available}.
Either allow the user to
configure the exact path to the package, \reviewed{distribute the program with the
dependent software, or download it during installation using your
package manager. If the executable requires internet access, check for that
early in execution.

@@

Exercise caution especially when transitioning across languages or using
separate executables, as they tend to be especially sensitive to operating
systems, environments, and locales.}

@@

\reviewed{To compile code, deploy applications, and automate other tasks,
programmers routinely use build tools like Make, Rake, Maven, Ant or MS Build.
These tools can also be used to manage runtime environments,
i.e.,
to check that the right versions of required packages are installed
and install or upgrade them if they are not.
As mentioned in Rule~5,
a package manager can mitigate some of the difficulties in software reuse.}

@@

Every package should come with a set of tests for users to run
after installation. Its purpose is not only to check that the software
is working correctly (although that is extremely helpful), but also to
ensure that it works at all. This test script can also serve as a
working example of how to run the software.

In order to be useful, \ruleminor{make the tests easy to find and run}.
Many build systems will also run unit tests if provided them at compile time.
\reviewed{For users, or
if the build system is not amenable to testing, 
provide a working script
in the project's root directory named \texttt{runtests.sh}
or something equally obvious. 
This lets new users build their analysis from a working script.
For example, with its distribution, 
HISAT2 includes a full set of very small files,
and a 'Getting Started with HISAT2' section in its manual 
that leads you through the entire data lifecycle~\cite{pertea2016}.}

@@

\reviewed{The usage message tells users what the program could do.
It is equally important for the program to tell users what it actually did.
Accordingly,
when the program starts, it should \ruleminor{echo all parameters and software
versions to standard out or a log file alongside the results}. This
feature supports greater reproducibility because any result can be
replicated with only the previous output files as reference.}
